---
title: "ANN"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: true
---

[← Back to Main Report](GroupFinalReport.html)

# Introduction 
This section trains an Artificial Neural Network to predict the history of mental illness outcome. The dummy-encoded and min-max normalized training and test sets are loaded, as the ANN requires all inputs to be numerical and on a comparable scale. The network is built using the neuralnet package with a two-hidden-layer architecture (5 neurons and 2 neurons respectively), and trained until convergence using backpropagation.


# import data

```{r}
train <- readRDS("train_dummy.rds")
test <- readRDS("test_dummy.rds")

str(train)
str(test)
```



```{r, cache=TRUE}
library(neuralnet)

ann_model <- neuralnet(
  formula = history_of_mental_illness_yes ~ .,
  data = train,

  hidden = c(5, 2),
  stepmax = 1e8,
  lifesign = "full")
plot(ann_model)

saveRDS(ann_model, file = "ann.rds")

```

```{r}
library(neuralnet)

# 1. Load your saved model and the dummy test data
ann_model <- readRDS("ann.rds")
test_dummy <- readRDS("test_dummy.rds")

# 2. Identify the target column index 
# (In your previous code, this was at ncol - 3)
target_idx <- ncol(test_dummy) - 3

# 3. Use compute() on the test data (excluding the target column)
# IMPORTANT: The columns must be in the EXACT same order as the training data
ann_results <- compute(ann_model, test_dummy[, -target_idx])

# 4. Extract the probabilities
# net.result is a matrix; we use as.vector() to make it a standard vector
p_ann_mental <- as.vector(ann_results$net.result)

# 5. Check the output
summary(p_ann_mental)
head(p_ann_mental)

# 6. Save the probability vector
saveRDS(p_ann_mental, "ann_prob_mental_illness.rds")
```

```{r}
library(caret)

# convert probabilities to class predictions
pred_class <- ifelse(p_ann_mental > 0.34, 1, 0)

# true labels
true_class <- test_dummy[, target_idx]

confusionMatrix(
  as.factor(pred_class),
  as.factor(true_class),
  positive = "1"
)
```
# Conclusion

A classification threshold of 0.34 is applied, consistent with the one selected for the GLM. The confusion matrix shows an accuracy of 64.3% and a Kappa of 0.07. While accuracy is slightly higher than the GLM, the Kappa is notably lower, indicating weaker overall agreement between predictions and true labels. Sensitivity drops to 24.5%, meaning the model struggles considerably to identify positive cases, while specificity is higher at 81.9%. The model performs better than random chance (Kappa > 0), but its predictive power on the positive class is limited.

