---
title: "Depression Model — Performance Visualization"
author: "Ye Tianci (Mike)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: flatly
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      fig.align = "center")
```

# Financial Parameters

```{r financial-params}
COST_FN  <- 763   # missed depressed patient
COST_FP  <- 246   # gave free therapy/resources to non-depressed
SAVING_TP <- 517  # savings from correctly identifying depressed patient
BENEFIT_TN <- 0   # correctly identified non-depressed

compute_financials <- function(tp, fp, fn, tn, model_name) {
  impact <- tp * SAVING_TP - fp * COST_FP - fn * COST_FN + tn * BENEFIT_TN
  precision <- ifelse((tp + fp) > 0, tp / (tp + fp), 0)
  data.frame(
    Model = model_name, TP = tp, FP = fp, FN = fn, TN = tn,
    Net_Impact = round(impact, 0),
    Precision = round(precision * 100, 2)
  )
}

find_best_threshold <- function(prob, actual, model_name, thresholds = seq(0.05, 0.95, by = 0.05)) {
  results <- data.frame()
  for (t in thresholds) {
    pred <- ifelse(prob > t, 1, 0)
    tp <- sum(pred == 1 & actual == "1")
    fp <- sum(pred == 1 & actual == "0")
    fn <- sum(pred == 0 & actual == "1")
    tn <- sum(pred == 0 & actual == "0")
    impact <- tp * SAVING_TP - fp * COST_FP - fn * COST_FN + tn * BENEFIT_TN
    results <- rbind(results, data.frame(Threshold = t, TP = tp, FP = fp, FN = fn, TN = tn, Net_Impact = impact))
  }
  results$Model <- model_name
  results
}
```

# Load Data and Predictions

```{r load-data}
library(caret)
library(C50)
library(randomForest)

test_dummy    <- readRDS("test_dummy.rds")
test_original <- readRDS("test_original.rds")

if ("mental_illness" %in% names(test_dummy)) {
  actual_labels <- as.factor(test_dummy$mental_illness)
} else if ("y" %in% names(test_dummy)) {
  actual_labels <- as.factor(test_dummy$y)
} else {
  actual_labels <- as.factor(test_dummy[, ncol(test_dummy)])
}
cat("Test set size:", length(actual_labels), "\n")
cat("Class distribution:\n")
print(prop.table(table(actual_labels)))

glm_prob  <- readRDS("glm_prob.rds")
knn_prob  <- readRDS("knn_prob_mental_illness.rds")
ann_prob  <- readRDS("ann_prob_mental_illness.rds")
c50_prob  <- readRDS("c50_prob_mental_illness.rds")
svm_prob  <- readRDS("best_svm_prob_mental_illness.rds")

svm_vanilla <- readRDS("svm_vanilla.rds")
svm_rbf     <- readRDS("svm_rbf.rds")
svm_poly    <- readRDS("svm_poly.rds")
svm_anova   <- readRDS("svm_anova.rds")
svm_bessel  <- readRDS("svm_bessel.rds")
svm_laplace <- readRDS("svm_laplace.rds")
svm_tanh    <- readRDS("svm_tanh.rds")

stack_test   <- readRDS("depression_stack_test.rds")
stack_train  <- readRDS("depression_stack_train.rds")
meta_tree    <- readRDS("depression_meta_model_tree.rds")
meta_rf      <- readRDS("depression_meta_model_rf.rds")

stack_test$actual  <- as.factor(stack_test$actual)
stack_train$actual <- as.factor(stack_train$actual)
```

# Level 1: Individual Model Performance

```{r l1-confusion}
flip_if_needed <- function(prob, actual) {
  pred_1 <- sum(prob > 0.5)
  if (pred_1 == 0 & mean(prob) > 0.3) {
    cat("  -> Flipping probabilities (likely stored as P(class=0))\n")
    return(1 - prob)
  }
  return(prob)
}

cat("Checking probability directions:\n")
cat("glm_prob range:", range(glm_prob), "\n")
glm_prob <- flip_if_needed(glm_prob, actual_labels)
cat("knn_prob range:", range(knn_prob), "\n")
knn_prob <- flip_if_needed(knn_prob, actual_labels)
cat("ann_prob range:", range(ann_prob), "\n")
ann_prob <- flip_if_needed(ann_prob, actual_labels)
cat("c50_prob range:", range(c50_prob), "\n")
c50_prob <- flip_if_needed(c50_prob, actual_labels)
cat("svm_prob range:", range(svm_prob), "\n")
svm_prob <- flip_if_needed(svm_prob, actual_labels)

glm_pred      <- factor(ifelse(glm_prob > 0.34, 1, 0), levels = levels(actual_labels))
knn_pred      <- factor(ifelse(knn_prob > 0.34, 1, 0), levels = levels(actual_labels))
ann_pred      <- factor(ifelse(ann_prob > 0.34, 1, 0), levels = levels(actual_labels))
c50_pred      <- factor(ifelse(c50_prob > 0.34, 1, 0), levels = levels(actual_labels))
svm_best_pred <- factor(ifelse(svm_prob > 0.34, 1, 0), levels = levels(actual_labels))

glm_cm <- confusionMatrix(glm_pred, actual_labels, positive = "1")
knn_cm <- confusionMatrix(knn_pred, actual_labels, positive = "1")
ann_cm <- confusionMatrix(ann_pred, actual_labels, positive = "1")
c50_cm <- confusionMatrix(c50_pred, actual_labels, positive = "1")
svm_cm <- confusionMatrix(svm_best_pred, actual_labels, positive = "1")
```

```{r l1-print}
cat("=== Logistic Regression ===\n"); print(glm_cm)
cat("\n=== KNN ===\n"); print(knn_cm)
cat("\n=== ANN ===\n"); print(ann_cm)
cat("\n=== Decision Tree (C5.0) ===\n"); print(c50_cm)
cat("\n=== Best SVM ===\n"); print(svm_cm)
```

# Threshold Optimization

Instead of using the default 0.5 threshold, we test thresholds from 0.05 to 0.95 in increments of 0.05 and pick the one that maximizes **net financial impact** for each model. Since missing a depressed patient costs \$800 vs \$251 for a false alarm, the optimal threshold is likely below 0.5.

```{r threshold-search}
glm_thresh <- find_best_threshold(glm_prob, actual_labels, "Logistic Regression")
knn_thresh <- find_best_threshold(knn_prob, actual_labels, "KNN")
ann_thresh <- find_best_threshold(ann_prob, actual_labels, "ANN")
c50_thresh <- find_best_threshold(c50_prob, actual_labels, "Decision Tree (C5.0)")
svm_thresh <- find_best_threshold(svm_prob, actual_labels, "Best SVM")

all_thresh <- rbind(glm_thresh, knn_thresh, ann_thresh, c50_thresh, svm_thresh)

best_per_model <- do.call(rbind, lapply(split(all_thresh, all_thresh$Model), function(df) {
  df[which.max(df$Net_Impact), ]
}))

cat("Best threshold per model (maximizing financial impact):\n")
print(best_per_model[, c("Model", "Threshold", "TP", "FP", "FN", "TN", "Net_Impact")])
```

```{r threshold-plot, fig.width=10, fig.height=6}
models_list <- list(glm_thresh, knn_thresh, ann_thresh, c50_thresh, svm_thresh)
model_colors <- c("#e74c3c", "#3498db", "#2ecc71", "#f39c12", "#9b59b6")
model_labels <- c("LogReg", "KNN", "ANN", "C5.0", "SVM")

plot(NULL, xlim = c(0.05, 0.95), ylim = range(all_thresh$Net_Impact),
     xlab = "Threshold", ylab = "Net Financial Impact ($)",
     main = "Financial Impact vs. Classification Threshold")
abline(h = 0, lty = 2, col = "gray40")
abline(v = 0.5, lty = 3, col = "gray60")
text(0.52, max(all_thresh$Net_Impact) * 0.95, "Default 0.5", col = "gray60", cex = 0.7)

for (i in seq_along(models_list)) {
  df <- models_list[[i]]
  lines(df$Threshold, df$Net_Impact, col = model_colors[i], lwd = 2)
  best_idx <- which.max(df$Net_Impact)
  points(df$Threshold[best_idx], df$Net_Impact[best_idx], col = model_colors[i], pch = 19, cex = 1.5)
}
legend("topright", legend = model_labels, col = model_colors, lwd = 2, bty = "n")
```

```{r threshold-table}
probs_list <- list(glm_prob, knn_prob, ann_prob, c50_prob, svm_prob)
prob_names <- c("Logistic Regression", "KNN", "ANN", "Decision Tree (C5.0)", "Best SVM")

default_impacts <- data.frame()
optimized_impacts <- data.frame()

for (i in seq_along(probs_list)) {
  prob <- probs_list[[i]]
  pred_def <- ifelse(prob > 0.5, 1, 0)
  tp <- sum(pred_def == 1 & actual_labels == "1"); fp <- sum(pred_def == 1 & actual_labels == "0")
  fn <- sum(pred_def == 0 & actual_labels == "1"); tn <- sum(pred_def == 0 & actual_labels == "0")
  impact_def <- tp * SAVING_TP - fp * COST_FP - fn * COST_FN

  best_row <- best_per_model[best_per_model$Model == prob_names[i], ]

  default_impacts <- rbind(default_impacts, data.frame(Model = prob_names[i], Threshold = 0.5, Net_Impact = impact_def))
  optimized_impacts <- rbind(optimized_impacts, data.frame(Model = prob_names[i], Threshold = best_row$Threshold, Net_Impact = best_row$Net_Impact))
}

thresh_comparison <- merge(default_impacts, optimized_impacts, by = "Model", suffixes = c("_Default", "_Optimized"))
thresh_comparison$Improvement <- thresh_comparison$Net_Impact_Optimized - thresh_comparison$Net_Impact_Default

knitr::kable(thresh_comparison[, c("Model", "Threshold_Default", "Net_Impact_Default",
                                    "Threshold_Optimized", "Net_Impact_Optimized", "Improvement")],
             align = "lccccc",
             col.names = c("Model", "Default Thresh", "Impact (Default)", "Best Thresh", "Impact (Optimized)", "$ Improvement"),
             caption = "Default (0.5) vs. Optimized Threshold — Financial Impact")
```

# Level 1 at Optimized Thresholds

```{r l1-optimized}
glm_best_t <- best_per_model[best_per_model$Model == "Logistic Regression", "Threshold"]
knn_best_t <- best_per_model[best_per_model$Model == "KNN", "Threshold"]
ann_best_t <- best_per_model[best_per_model$Model == "ANN", "Threshold"]
c50_best_t <- best_per_model[best_per_model$Model == "Decision Tree (C5.0)", "Threshold"]
svm_best_t <- best_per_model[best_per_model$Model == "Best SVM", "Threshold"]

glm_pred_opt <- factor(ifelse(glm_prob > glm_best_t, 1, 0), levels = levels(actual_labels))
knn_pred_opt <- factor(ifelse(knn_prob > knn_best_t, 1, 0), levels = levels(actual_labels))
ann_pred_opt <- factor(ifelse(ann_prob > ann_best_t, 1, 0), levels = levels(actual_labels))
c50_pred_opt <- factor(ifelse(c50_prob > c50_best_t, 1, 0), levels = levels(actual_labels))
svm_pred_opt <- factor(ifelse(svm_prob > svm_best_t, 1, 0), levels = levels(actual_labels))

glm_cm_opt <- confusionMatrix(glm_pred_opt, actual_labels, positive = "1")
knn_cm_opt <- confusionMatrix(knn_pred_opt, actual_labels, positive = "1")
ann_cm_opt <- confusionMatrix(ann_pred_opt, actual_labels, positive = "1")
c50_cm_opt <- confusionMatrix(c50_pred_opt, actual_labels, positive = "1")
svm_cm_opt <- confusionMatrix(svm_pred_opt, actual_labels, positive = "1")

model_names_opt <- c("LogReg (opt)", "KNN (opt)", "ANN (opt)", "C5.0 (opt)", "SVM (opt)")
cms_opt <- list(glm_cm_opt, knn_cm_opt, ann_cm_opt, c50_cm_opt, svm_cm_opt)

comparison_opt <- data.frame(
  Model = model_names_opt,
  Threshold = c(glm_best_t, knn_best_t, ann_best_t, c50_best_t, svm_best_t),
  Accuracy = round(sapply(cms_opt, function(x) x$overall["Accuracy"]), 4),
  Kappa = round(sapply(cms_opt, function(x) x$overall["Kappa"]), 4),
  Sensitivity = round(sapply(cms_opt, function(x) x$byClass["Sensitivity"]), 4),
  Specificity = round(sapply(cms_opt, function(x) x$byClass["Specificity"]), 4)
)

knitr::kable(comparison_opt, align = "lccccc", caption = "Level 1 Models at Optimized Thresholds")
```

# Level 2: Stacked Model Performance

```{r l2-predictions}
test_features <- stack_test[, -which(names(stack_test) == "actual")]
test_labels_stack <- stack_test$actual

meta_tree_pred <- predict(meta_tree, newdata = test_features)
meta_rf_pred   <- predict(meta_rf, newdata = test_features)

meta_tree_cm <- confusionMatrix(meta_tree_pred, test_labels_stack, positive = "1")
meta_rf_cm   <- confusionMatrix(meta_rf_pred, test_labels_stack, positive = "1")

cat("=== Stacked Model (C5.0 Meta-Tree) ===\n"); print(meta_tree_cm)
cat("\n=== Stacked Model (Random Forest) ===\n"); print(meta_rf_cm)
```

# Full Comparison

```{r full-compare}
model_names_def <- c("Logistic Regression", "KNN", "ANN", "Decision Tree (C5.0)", "Best SVM")
cms_def <- list(glm_cm, knn_cm, ann_cm, c50_cm, svm_cm)

comp_def <- data.frame(
  Model = model_names_def,
  Accuracy = round(sapply(cms_def, function(x) x$overall["Accuracy"]), 4),
  Kappa = round(sapply(cms_def, function(x) x$overall["Kappa"]), 4),
  Sensitivity = round(sapply(cms_def, function(x) x$byClass["Sensitivity"]), 4),
  Specificity = round(sapply(cms_def, function(x) x$byClass["Specificity"]), 4)
)
comp_stack <- data.frame(
  Model = c("Stacked (C5.0)", "Stacked (RF)"),
  Accuracy = round(c(meta_tree_cm$overall["Accuracy"], meta_rf_cm$overall["Accuracy"]), 4),
  Kappa = round(c(meta_tree_cm$overall["Kappa"], meta_rf_cm$overall["Kappa"]), 4),
  Sensitivity = round(c(meta_tree_cm$byClass["Sensitivity"], meta_rf_cm$byClass["Sensitivity"]), 4),
  Specificity = round(c(meta_tree_cm$byClass["Specificity"], meta_rf_cm$byClass["Specificity"]), 4)
)

comparison_all <- rbind(comp_def, comparison_opt[, -2], comp_stack)
knitr::kable(comparison_all, align = "lcccc", caption = "All Models — Full Statistical Comparison")
```

```{r full-kappa, fig.width=12, fig.height=5}
par(mar = c(12, 5, 3, 2))
kv <- comparison_all$Kappa; kv[is.na(kv)] <- 0
sn <- c("LogReg", "KNN", "ANN", "C5.0", "SVM",
        "LogReg*", "KNN*", "ANN*", "C5.0*", "SVM*",
        "Stack-Tree", "Stack-RF")
names(kv) <- sn
kc <- c(rep("#3498db",5), rep("#f39c12",5), rep("#e74c3c",2))
kc[which.max(kv)] <- "#2ecc71"
barplot(kv, col=kc, main="Kappa (* = optimized threshold)", ylab="Kappa", las=2, cex.names=0.7,
        ylim=c(0, max(kv,na.rm=TRUE)*1.3))
legend("topright", legend=c("Default","Optimized","Stacked"), fill=c("#3498db","#f39c12","#e74c3c"), bty="n", cex=0.8)
```

```{r full-sens-spec, fig.width=12, fig.height=5}
par(mfrow=c(1,2), mar=c(12,5,3,2))
sv <- comparison_all$Sensitivity; sv[is.na(sv)] <- 0; names(sv) <- sn
sc <- c(rep("#3498db",5), rep("#f39c12",5), rep("#e74c3c",2)); sc[which.max(sv)] <- "#2ecc71"
barplot(sv, col=sc, main="Sensitivity (* = optimized)", ylab="Sensitivity", las=2, cex.names=0.65, ylim=c(0,1))

spv <- comparison_all$Specificity; spv[is.na(spv)] <- 0; names(spv) <- sn
spc <- c(rep("#3498db",5), rep("#f39c12",5), rep("#e74c3c",2)); spc[which.max(spv)] <- "#2ecc71"
barplot(spv, col=spc, main="Specificity (* = optimized)", ylab="Specificity", las=2, cex.names=0.65, ylim=c(0,1))
```

# Financial Comparison

```{r fin-compare}
all_cms <- c(cms_def, cms_opt, list(meta_tree_cm, meta_rf_cm))
all_names <- c(model_names_def, model_names_opt, "Stacked (C5.0)", "Stacked (RF)")

fin_results <- data.frame()
for (i in seq_along(all_cms)) {
  cm <- all_cms[[i]]$table
  tp <- cm["1","1"]; fp <- cm["1","0"]; fn <- cm["0","1"]; tn <- cm["0","0"]
  fin_results <- rbind(fin_results, compute_financials(tp, fp, fn, tn, all_names[i]))
}

n_pos <- sum(actual_labels == "1"); n_neg <- sum(actual_labels == "0")
baseline <- compute_financials(0, 0, n_pos, n_neg, "Baseline")
baseline_impact <- baseline$Net_Impact

fin_results$Net_Impact <- fin_results$Net_Impact - baseline_impact
baseline$Net_Impact <- 0
fin_results <- rbind(baseline, fin_results)

knitr::kable(fin_results[, c("Model", "TP", "FP", "FN", "TN", "Net_Impact", "Precision")],
             align = "lcccccc", caption = "Financial Impact vs. Baseline (Relative)")
```

```{r fin-plot, fig.width=12, fig.height=6}
par(mar = c(12, 5, 3, 2))
iv <- fin_results$Net_Impact
fn <- c("Baseline", sn)
names(iv) <- fn
fc <- c("gray60", rep("#3498db",5), rep("#f39c12",5), rep("#e74c3c",2))
fc[which.max(iv)] <- "#2ecc71"
barplot(iv, col=fc, main="Net Financial Impact vs. Baseline (* = optimized threshold)",
        ylab="Net Impact ($)", las=2, cex.names=0.6)
abline(h=0, lty=2, col="gray40")
legend("topright", legend=c("Default","Optimized","Stacked"), fill=c("#3498db","#f39c12","#e74c3c"), bty="n", cex=0.8)
```

# SVM Kernel Comparison

```{r svm-kernels}
svm_models <- list(svm_vanilla, svm_rbf, svm_poly, svm_anova, svm_bessel, svm_laplace, svm_tanh)
svm_knames <- c("Vanilla", "RBF", "Polynomial", "ANOVA", "Bessel", "Laplace", "Tanh")
svm_compare <- data.frame()
for (i in seq_along(svm_models)) {
  m <- svm_models[[i]]
  svm_compare <- rbind(svm_compare, data.frame(
    Kernel = svm_knames[i], Type = class(m)[1],
    Error = tryCatch(m@error, error = function(e) NA),
    Cross_Validation = tryCatch(m@cross, error = function(e) NA)))
}
knitr::kable(svm_compare, align = "lccc", caption = "SVM Kernel Comparison")
```

# Summary

```{r summary-vars}
best_fin_name <- fin_results$Model[which.max(fin_results$Net_Impact)]
best_fin_val <- max(fin_results$Net_Impact)
best_kappa_name <- comparison_all$Model[which.max(comparison_all$Kappa)]
best_kappa_val <- max(comparison_all$Kappa)
best_sens_name <- comparison_all$Model[which.max(comparison_all$Sensitivity)]
best_sens_val <- max(comparison_all$Sensitivity)
```

The best model by Kappa is **`r best_kappa_name`** (`r round(best_kappa_val, 4)`), the best by sensitivity is **`r best_sens_name`** (`r round(best_sens_val, 4)`), and the best by financial impact is **`r best_fin_name`** at \$`r format(best_fin_val, big.mark=",")` relative to baseline.

Threshold optimization makes a significant difference — the default 0.5 cutoff is suboptimal for every model because the cost of missing a depressed patient (\$800) far exceeds the cost of a false alarm (\$251). By lowering the threshold, models flag more patients as potentially depressed, accepting more false positives to catch more true cases — which is the financially correct tradeoff.
