---
title: "DT2"
author: "Daniel Myung"
date: "2026-04-13"
output: html_document
---

[← Back to Main Report](GroupFinalReport.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this section, we apply two tree-based classifiers to predict whether an individual has a history of mental illness: a C5.0 Decision Tree and a Random Forest. 

The C5.0 decision tree recursively partitions the data based on the feature that best separates the classes at each node, producing an interpretable set of if-else rules. The Random Forest extends this idea by building an ensemble of 500 decision trees, each trained on a random bootstrap sample of the data and considering only a random subset of features at each split. 

For the Random Forest, a custom classification threshold of 0.34 is also explored to improve sensitivity toward the minority class. 

```{r, cache= TRUE}
library(C50)
library(caret)
set.seed(12345)

# Using the original data as requested
train <- readRDS("train_original.rds")
test <- readRDS("test_original.rds")

# C5.0 requires the target to be a factor
train$History.of.Mental.Illness <- as.factor(train$History.of.Mental.Illness)
test$History.of.Mental.Illness <- as.factor(test$History.of.Mental.Illness)

# Define the features (X) and the target (y)
# We exclude the target column from the training features
features <- train[, names(train) != "History.of.Mental.Illness"]
target   <- train$History.of.Mental.Illness

# Create the C5.0 model

c50_model <- C5.0(features, target, trials = 1)

# If you want to be explicit about the package:
# c50_model <- C50::C5.0(features, target, trials = 1)

summary(c50_model)

# Note: C5.0 trees can get very large, plot may be dense
plot(c50_model)

# Predict Class
c50_pred <- predict(c50_model, test, type = "class")

# Predict Probabilities (for ROC curves later)
c50_prob <- predict(c50_model, test, type = "prob")

head(c50_pred)
head(c50_prob)

confusionMatrix(c50_pred, test$History.of.Mental.Illness, positive = "Yes")

saveRDS(c50_model, file = "c50_model.rds")
saveRDS(c50_pred, file = "c50_pred.rds")
# Saving the 'Yes' probability vector specifically for consistency with your other models
saveRDS(c50_prob[, "Yes"], file = "c50_prob_mental_illness.rds")
```

The C5.0 tree produces a degenerate result: it predicts "No" for every single observation, achieving an accuracy of 69.2% that exactly matches the No Information Rate. The Kappa of 0 confirms the model has learned nothing beyond the majority class baseline, with sensitivity of 0% and specificity of 100%. The tree collapsed to a single root node, meaning it found no split informative enough to justify branching: this is a direct consequence of the class imbalance overwhelming the algorithm.

```{r, cache= TRUE}
library(randomForest)
library(caret)

# 1. LOAD DATA
# Using the original data as requested for tree-based models
train <- readRDS("train_original.rds")
test  <- readRDS("test_original.rds")

# 2. PRE-PROCESSING
# Ensure target variable is a factor for classification
train$History.of.Mental.Illness <- as.factor(train$History.of.Mental.Illness)
test$History.of.Mental.Illness  <- as.factor(test$History.of.Mental.Illness)

# Define the features (X) and the target (y)
features <- train[, names(train) != "History.of.Mental.Illness"]
target   <- train$History.of.Mental.Illness

# 3. TRAIN RANDOM FOREST
# ntree: number of trees (500 is standard)
# importance: TRUE allows us to see which variables matter most
set.seed(123) # For reproducibility
rf_model <- randomForest(features, target, ntree = 500, importance = TRUE)

# View model summary and variable importance
print(rf_model)
importance(rf_model)
varImpPlot(rf_model)

# 4. PREDICTIONS
# Ensure test features match training features
test_features <- test[, names(test) != "History.of.Mental.Illness"]

# Predict Class (Default 0.5 threshold)
rf_pred <- predict(rf_model, test_features, type = "class")

# Predict Probabilities
rf_prob <- predict(rf_model, test_features, type = "prob")

# 5. EVALUATION
# Using the 0.5 threshold first
cat("\n--- RANDOM FOREST (Default 0.5 Threshold) ---\n")
confusionMatrix(rf_pred, test$History.of.Mental.Illness, positive = "Yes")

# 6. OPTIONAL: APPLY CUSTOM THRESHOLD (0.34)
# Since you mentioned the 0.5 threshold predicts "All No", this 0.34 shift 
# will make the model more sensitive.
rf_pred_34 <- factor(ifelse(rf_prob[, "Yes"] > 0.34, "Yes", "No"), levels = c("No", "Yes"))

cat("\n--- RANDOM FOREST (Adjusted 0.34 Threshold) ---\n")
confusionMatrix(rf_pred_34, test$History.of.Mental.Illness, positive = "Yes")

# 7. SAVE OUTPUTS
saveRDS(rf_model, file = "rf_model.rds")
saveRDS(rf_pred_34, file = "rf_pred.rds") # Saving the sensitive version for the council
saveRDS(rf_prob[, "Yes"], file = "rf_prob_mental_illness.rds")

# Preview results
head(rf_prob)
```

At the default 0.5 threshold, the Random Forest suffers from the same majority-class bias, with a Kappa of just 0.018 and sensitivity of only 3.8%. Lowering the threshold to 0.34 meaningfully improves performance: Kappa rises to 0.095 and sensitivity reaches 35%, at the cost of reduced specificity (74.3%). This represents a more balanced trade-off and is the more useful configuration for this task.

At this point, the GLM remains the best overall model with a Kappa of 0.143 and the most balanced sensitivity (43.8%) and specificity (70.8%). The Random Forest at the adjusted threshold is second, followed by KNN, while the C5.0 tree fails entirely.