---
title: "GLM - Depression (History of Mental Illness)"
output: html_document
---

# Import data

```{r}
train <- readRDS("train_original.rds")
test  <- readRDS("test_original.rds")
```


# Exploratory check

```{r}
str(train)
table(train$History.of.Mental.Illness)
prop.table(table(train$History.of.Mental.Illness))
```


# Fit GLM (logistic regression)

```{r cache=TRUE}
glm_model <- glm(History.of.Mental.Illness ~ .,
                 data   = train,
                 family = binomial)
saveRDS(glm_model, file = "glm_model.rds")
summary(glm_model)
```


# Odds Ratios
The GLM estimates coefficients in log-odds, which are difficult to interpret directly. By applying the exponential function (exp()), we convert them into Odds Ratios (OR), which express how much the odds of having a history of mental illness change for each predictor. An OR greater than 1 indicates that the variable increases the probability of "Yes", while an OR less than 1 indicates a decrease. An OR equal to 1 means no effect. The 95% confidence interval is also reported: if it does not include 1, the effect is statistically significant.

```{r cache=TRUE}
exp(cbind(OR = coef(glm_model), confint(glm_model)))
```


# Predictions on test set

```{r}
glm_prob <- predict(glm_model, newdata = test, type = "response")

glm_pred <- ifelse(glm_prob > 0.34, "Yes", "No")
glm_pred <- factor(glm_pred, levels = levels(test$History.of.Mental.Illness))

saveRDS(glm_prob,  file = "glm_prob.rds")
saveRDS(glm_pred, file = "glm_pred.rds")
```


# Confusion Matrix

```{r}
library(caret)

confusionMatrix(glm_pred, test$History.of.Mental.Illness, positive = "Yes")
```



