---
title: "GLM - Depression (History of Mental Illness)"
output: html_document
---

# Import data

```{r}
train <- readRDS("train_original.rds")
test  <- readRDS("test_originial.rds")
```


# Exploratory check

```{r}
str(train)
table(train$History.of.Mental.Illness)
prop.table(table(train$History.of.Mental.Illness))
```


# Fit GLM (logistic regression)

```{r cache=TRUE}
glm_model <- glm(History.of.Mental.Illness ~ .,
                 data   = train,
                 family = binomial)
saveRDS(glm_model, file = "glm_model.rds")
summary(glm_model)
```


# Odds Ratios
The GLM estimates coefficients in log-odds, which are difficult to interpret directly. By applying the exponential function (exp()), we convert them into Odds Ratios (OR), which express how much the odds of having a history of mental illness change for each predictor. An OR greater than 1 indicates that the variable increases the probability of "Yes", while an OR less than 1 indicates a decrease. An OR equal to 1 means no effect. The 95% confidence interval is also reported: if it does not include 1, the effect is statistically significant.

```{r cache=TRUE}
exp(cbind(OR = coef(glm_model), confint(glm_model)))
```


# Predictions on test set

```{r}
glm_prob <- predict(glm_model, newdata = test, type = "response")

glm_pred <- ifelse(glm_prob > 0.28, "Yes", "No")
glm_pred <- factor(glm_pred, levels = levels(test$History.of.Mental.Illness))

saveRDS(glm_prob,  file = "glm_prob.rds")
```


# Confusion Matrix

```{r}
library(caret)

confusionMatrix(glm_pred, test$History.of.Mental.Illness, positive = "Yes")
```


# ROC Curve and AUC
The ROC curve (Receiver Operating Characteristic) is a graphical tool that evaluates the performance of a binary classifier across all possible classification thresholds. It plots the True Positive Rate (Sensitivity) on the y-axis against the False Positive Rate (1 - Specificity) on the x-axis. A model that predicts randomly would follow the diagonal, while a perfect model would reach the top-left corner. The AUC (Area Under the Curve) summarizes the ROC curve into a single number between 0 and 1. As a general guideline: AUC around 0.5 means the model performs no better than random chance, between 0.7 and 0.8 is considered acceptable, between 0.8 and 0.9 is good, and above 0.9 is excellent. In a medical context, the ROC curve is particularly useful because it allows us to choose the classification threshold that best balances sensitivity and specificity based on the specific goal — in our case, prioritizing sensitivity to avoid missing true cases.

```{r}
library(pROC)

roc_obj <- roc(test$History.of.Mental.Illness, glm_prob, levels = c("No", "Yes"))

plot(roc_obj,
     col  = "steelblue",
     main = paste("ROC Curve - AUC:", round(auc(roc_obj), 4)))

auc(roc_obj)
```
