---
title: "GroupFinalReport"
output:
  html_document:
    css: ross-rmarkdown.css
    theme: flatly
    toc: true
    toc_float:
      collapsed: true
    code_folding: hide
    includes:
      before_body: ross-header.html
---
| ![Riccardo Ferrari](https://drive.google.com/uc?export=view&id=1DtvCv0cgOIgukiAOCnaYlYNkAS5bPZ0C "Riccardo Ferrari") | ![Mike Ye](https://drive.google.com/uc?export=view&id=1MGh2iKI855FLUBuEJzPAOURbKulWka9r "Mike Ye") | ![Daniel Myung](https://drive.google.com/uc?export=view&id=1oX3kGp_eOsgvzug_oAhZMJBVH7htjpfW) | ![Steven Liu](https://drive.google.com/uc?export=view&id=10eQB_7168yJiNFwlv98QrgtHAe1rZQov "Steven Liu") | ![Andrea Demelas](https://drive.google.com/uc?export=view&id=1sfs8yje5_bIvdE-VDuWeBrWkE1UmgzFm) |
|---|---|---|---|---|
| **[Riccardo Ferrari](https://www.linkedin.com/in/riccardo-ferrari-2929b722a/)**<br>Financial Analyst | **[Mike Ye](https://www.linkedin.com/in/mike-ye-460923270/)**<br>Team Lead | **[Daniel Myung]()**<br>Data Scientist | **[Steven Liu](https://www.linkedin.com/in/steven-liu-554652244/)**<br>Communication Specialist | **[Andrea Demelas](https://www.linkedin.com/in/andreademelas/)**<br>Business Analyst |

# Executive Summary 

414 Insurance's dataset with depression data can reveal key characteristics as early mental illness predictors. Being able to quickly identify at-risk patients is an important priority for 414 Insurance, as it leads to preventions of the most extreme mental illness scenarios. 

We use data-forward methodologies to assess mental illness patterns, recommending a targeted approach to predict onset of mental illness. Our approach utilizes supervised and unsupervised algorithmic learning models to find target characteristics. We demonstrate our work with each type of algorithm to propose the most effective strategy for identifying and preventing onset of mental illness. 

# Background

Depression--characterized by symptoms ranging from persistent low mood, loss of interest, sucidial thoughts-- is a common, clinically significant, and expensive medical condition for insurers. According to a Gallup poll, 3 in 10 US adults have been diagnosed with depression at some point in their lives ([Gallup](https://news.gallup.com/poll/505745/depression-rates-reach-new-highs.aspx)). In colleges across the US, 37% of the student population reported moderate to severe depressive symptoms in 2025 ([University of Michigan](https://sph.umich.edu/news/2025posts/college-student-mental-health-third-consecutive-year-improvement.html)). 

While there's a perception of depression as a benign issue, depression can cause difficulty maintaining relationships, work, or self-care. This causes it to compound into other chronic conditions, which increases total medical spending. In fact, 414 Insurance sees a 73% difference--a difference of $763-- between insurees with and without depression ([Health System Tracker](https://www.healthsystemtracker.org/brief/privately-insured-people-with-depression-and-anxiety-face-high-out-of-pocket-costs/)). 

Therefore, identifying high-risk members earlier opens an opportunity to target interventions before costs escalate ([Greenberg et al](https://pmc.ncbi.nlm.nih.gov/articles/PMC10499687/)).

# Technical Analysis & Model Modules

The following links provide the detailed technical breakdown of our data processing and modeling pipeline.

<div id="Cleaning">

# 1. Data Preparation and Exploration
## Main
The first phase of the analysis focuses on preparing the data for modeling. The raw dataset is imported from a CSV file and a random 10% sample is drawn to reduce computational burden. The Name column is removed as it carries no predictive value, and all character variables are coerced to factors. A preliminary inspection of the data structure and summary statistics is then performed. The cleaned dataset is subsequently saved as an RDS file for reproducibility.

* [Main Data Cleaning](Main.html)

## Initial Training and Testing Split
In the second file, the cleaned data is split into training and test sets following two parallel pipelines. The first preserves the original factor structure for tree-based models. The second converts the data into a fully dummy-encoded and min-max normalized format, suitable for distance- and gradient-based models such as KNN, ANN, and SVM. Both pipelines use a 50/50 split with the same random seed to ensure comparability across models.

* [Initial Training and Testing Split](trainingtestingsplit.html)
</div>

# 2. Supervised Learning Models
* [Generalized Linear Model (GLM) for Depression](glm_depression.html)

The GLM is our best individual model, with a Kappa of 0.143 and the most balanced sensitivity (43.8%) and specificity (70.8%). It correctly identifies a meaningful share of both positive and negative cases, making it the most reliable classifier among the individual models we tested. We believe it performs best because the predictors in this dataset relate to mental illness risk in a relatively straightforward way, and the GLM is well-suited to capture that. It also handles the class imbalance better than our other models without needing threshold adjustments.

* [K-Nearest Neighbors (kNN) Analysis](knn_2.html)

KNN performs poorly, with a Kappa of just 0.026 and a sensitivity of only 9.5%, meaning it almost entirely fails to identify individuals with a history of mental illness. On the surface its accuracy looks reasonable at 67%, but this is misleading: a model that always predicts "No" would score similarly. We believe the main issue is the class imbalance: with roughly 69% of cases being "No," the model's nearest neighbors are almost always dominated by the majority class, pushing nearly every prediction toward "No" regardless of the individual's actual characteristics.

* [Decision Tree and Random Forest Results](DT2.html)

The C5.0 decision tree degenerates entirely, predicting "No" for every single observation and achieving a Kappa of 0. The Random Forest does better, but only after we lower the classification threshold from 0.5 to 0.34, which brings sensitivity up to 35% and Kappa to 0.095. We believe both models struggle for the same reason. The class imbalance is strong enough that neither algorithm finds it worthwhile to split toward the minority class without explicit correction. The Random Forest's ensemble approach gives it a slight edge, but it still falls well short of the GLM.

* [Support Vector Machine (SVM) Classification](SVM.html)

SVM proved to be the most inconsistent model in our analysis. Most of the kernels we tested (including the commonly used RBF and linear kernels) failed completely, predicting "No" for every observation. The splinedot kernel performed best with a Kappa of ~0.05 and sensitivity of ~0.30, and applying a lower threshold to the RBF model pushed sensitivity higher but introduced too many false positives to be practical. 

* [Artificial Neural Network (ANN) & Confusion Matrix](ANN-with-conf-matrix.html)

# 3. Ensemble Methods & Final Predictions

## Overview of Stacked Model Approach

The stacked model belongs to the family of stacked generalization (ensemble techniques) where a super‑learner (here, a decision tree) learns to combine base‑learner predictions in a way that can outperform any single model, strengthened by incorporating a cost matrix. 

In practice, each base model first generates a prediction vector on the test set, and then these predictions are coalesced into a single data frame along with the true response to create a new, smaller feature space for the second‑level model.  

### What is a cost matrix? 

In addition to combining the predictions of the five base models, the stacked model incorporates a **cost matrix** into the second‑level decision tree. 

In our context, falsely predicting that a non-depressed client is depressed (a false positive) may be an overuse of mental resources, while missing a depressed client (a false negative) creates large costs for 414 Insurance down the road. 

By assigning a penalty each misclassification, a cost matrix helps the decision tree optimize for a cost‑sensitive error tradeoff. 

## Kappa Analysis: Stacked Model

The stacked model achieved a relatively strong Kappa (9.81). 

It seemed like the stacked model mainly pulled from the ANN and KNN individual models. We hypothesize that since depression can exhibit complex patterns, the ANN was able to distinguish nonlinear relationships in the dataset.  The KNN contributes additional correlative relationships, creating clusters of similar depression groups that ANN  piggybacks off of.  This extra information lets the stacker better separate true positives from false positives, which shows up as a large jump in kappa. 


* [Combined Probabilities](CombinedProbs.html)

Here we combined the individual probability predictions from each of the models. Later on we use decision trees as the second layer of our stacked model. By taking the individual probabilities from each of the models and combining them, we create a dataset that allows the second layer to learn from. 

* [Secondary Data Partitioning](SecondSplitData.html)

Here we took the combined probabilities from the models, and created a 70 30 split to train and test our stacked models on. 

* [Stacked Model Architecture](StackedModel.html)
* [Final Stacked Model Predictions](StackModelPredictions.html)


# 4. Interactive Tools
Our ensemble model is hosted as a web application for real-time risk assessment.

<div style="text-align: center; margin: 20px 0;">
  <a href="https://j33c4y-daniel-myung.shinyapps.io/Interactive/" 
     target="_blank" 
     class="btn btn-success btn-lg" 
     style="padding: 15px 30px; font-weight: bold;">
     Launch Interactive Prediction Portal
  </a>
</div>

### 4. Financial analysis 
Although the statistical performance of our models are significant, the true measure of their efficacy is in their financial performance. More specifically, we define this to be the amount of money the models will save 414 insurance compared to the base line of labeling all patients as non-depressed. 

To ensure that this analysis is as accurate as possible, we have created the following cost matrix from real world data:

* **False Negative** (the model missed a depressed patient)............................**Cost**: \$800
* **False Positive** (incorrectly labeled a non-depressed patient as depressed)....**Cost**: \$251
* **True Negative** (correctly identified a non-depressed patient).....................**Cost**: \$0
* **True Positive** (correcting predicting a depressed patient).........................**Savings**: \$549

More specically, we found a difference of \$763 between non-depressed patient spending and depressed patient spending ([Health System Tracking](https://www.healthsystemtracker.org/brief/privately-insured-people-with-depression-and-anxiety-face-high-out-of-pocket-costs/#Annual%20out-of-pocket%C2%A0spending%20for%20privately-insured%20enrollees%20treated%20for%20depression%20and/or%20anxiety,%20compared%20to%20enrollees%20with%20no%20mental%20health%20diagnoses,%202021)). We rounded to $800 to account for economic burden. We found the savings from therapy and costs of depression from [Statista](https://www.statista.com/statistics/1481121/depression-cost-and-potential-savings-with-pat-in-the-us-and-europe/) (249,000 and 366,600). 

We created a savings-to-cost ratio using the Statista data. Solving according to the cost difference from Health System Tracking, we found a savings of $549. We use this savings to solve for the missing cost of providing mental health resources (800 - x = 549), which is $251.


Previously, we have set the classification thresholds (the lowest probability required to label an individual as "depressed") for each model to maximize their statistical performance, specially the kappa value. However, in a real world use case, the target we should optimize for changes to finiacial impact instead. Therefore, we have re-calibrated the classification thresholds for all models except for the stack model to maximize their performance, and our results are shown in the figures below. Note that since the stack models are predicting a smaller portion of the dataset, their financial impacts needed to be scaled accordingly to make this comparison useful. 

```{r dependencies, cache=TRUE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(caret)
library(C50)
library(randomForest)

# --- Financial parameters & helper ---
COST_FN  <- 800
COST_FP  <- 251
SAVING_TP <- 549
BENEFIT_TN <- 0

compute_financials <- function(tp, fp, fn, tn, model_name) {
  impact <- tp * SAVING_TP - fp * COST_FP - fn * COST_FN + tn * BENEFIT_TN
  precision <- ifelse((tp + fp) > 0, tp / (tp + fp), 0)
  data.frame(
    Model = model_name, TP = tp, FP = fp, FN = fn, TN = tn,
    Net_Impact = round(impact, 0),
    Precision = round(precision * 100, 2)
  )
}

find_best_threshold <- function(prob, actual, model_name, thresholds = seq(0.05, 0.95, by = 0.05)) {
  results <- data.frame()
  for (t in thresholds) {
    pred <- ifelse(prob > t, 1, 0)
    tp <- sum(pred == 1 & actual == "1")
    fp <- sum(pred == 1 & actual == "0")
    fn <- sum(pred == 0 & actual == "1")
    tn <- sum(pred == 0 & actual == "0")
    impact <- tp * SAVING_TP - fp * COST_FP - fn * COST_FN + tn * BENEFIT_TN
    results <- rbind(results, data.frame(Threshold = t, TP = tp, FP = fp, FN = fn, TN = tn, Net_Impact = impact))
  }
  results$Model <- model_name
  results
}

# --- Load data ---
test_dummy <- readRDS("test_dummy.rds")
if ("mental_illness" %in% names(test_dummy)) {
  actual_labels <- as.factor(test_dummy$mental_illness)
} else if ("y" %in% names(test_dummy)) {
  actual_labels <- as.factor(test_dummy$y)
} else {
  actual_labels <- as.factor(test_dummy[, ncol(test_dummy)])
}

glm_prob <- readRDS("glm_prob.rds")
knn_prob <- readRDS("knn_prob_mental_illness.rds")
ann_prob <- readRDS("ann_prob_mental_illness.rds")
c50_prob <- readRDS("c50_prob_mental_illness.rds")
svm_prob <- readRDS("best_svm_prob_mental_illness.rds")

stack_test <- readRDS("depression_stack_test.rds")
meta_tree  <- readRDS("depression_meta_model_tree.rds")
meta_rf    <- readRDS("depression_meta_model_rf.rds")
stack_test$actual <- as.factor(stack_test$actual)

# --- Flip probabilities if needed ---
flip_if_needed <- function(prob, actual) {
  if (sum(prob > 0.5) == 0 & mean(prob) > 0.3) return(1 - prob)
  return(prob)
}
glm_prob <- flip_if_needed(glm_prob, actual_labels)
knn_prob <- flip_if_needed(knn_prob, actual_labels)
ann_prob <- flip_if_needed(ann_prob, actual_labels)
c50_prob <- flip_if_needed(c50_prob, actual_labels)
svm_prob <- flip_if_needed(svm_prob, actual_labels)

# --- Default-threshold confusion matrices ---
model_names_def <- c("Logistic Regression", "KNN", "ANN", "Decision Tree (C5.0)", "Best SVM")
probs_list <- list(glm_prob, knn_prob, ann_prob, c50_prob, svm_prob)

cms_def <- lapply(probs_list, function(p) {
  pred <- factor(ifelse(p > 0.34, 1, 0), levels = levels(actual_labels))
  confusionMatrix(pred, actual_labels, positive = "1")
})

# --- Optimized-threshold confusion matrices ---
all_thresh <- do.call(rbind, lapply(seq_along(probs_list), function(i)
  find_best_threshold(probs_list[[i]], actual_labels, model_names_def[i])
))
best_per_model <- do.call(rbind, lapply(split(all_thresh, all_thresh$Model), function(df)
  df[which.max(df$Net_Impact), ]
))

model_names_opt <- c("LogReg (opt)", "KNN (opt)", "ANN (opt)", "C5.0 (opt)", "SVM (opt)")
best_thresholds <- sapply(model_names_def, function(m) best_per_model[best_per_model$Model == m, "Threshold"])

cms_opt <- lapply(seq_along(probs_list), function(i) {
  pred <- factor(ifelse(probs_list[[i]] > best_thresholds[i], 1, 0), levels = levels(actual_labels))
  confusionMatrix(pred, actual_labels, positive = "1")
})

# --- Stacked model confusion matrices ---
test_features <- stack_test[, -which(names(stack_test) == "actual")]
meta_tree_cm <- confusionMatrix(predict(meta_tree, newdata = test_features), stack_test$actual, positive = "1")
meta_rf_cm   <- confusionMatrix(predict(meta_rf, newdata = test_features), stack_test$actual, positive = "1")

# --- Short names for bar labels ---
sn <- c("LogReg", "KNN", "ANN", "C5.0", "SVM",
        "LogReg*", "KNN*", "ANN*", "C5.0*", "SVM*",
        "Stack-Tree", "Stack-RF")

# ====== FINANCIAL COMPARISON ======

all_cms <- c(cms_def, cms_opt, list(meta_tree_cm, meta_rf_cm))
all_names <- c(model_names_def, model_names_opt, "Stacked (C5.0)", "Stacked (RF)")

fin_results <- data.frame()
for (i in seq_along(all_cms)) {
  cm <- all_cms[[i]]$table
  tp <- cm["1","1"]; fp <- cm["1","0"]; fn <- cm["0","1"]; tn <- cm["0","0"]
  fin_results <- rbind(fin_results, compute_financials(tp, fp, fn, tn, all_names[i]))
}

n_pos <- sum(actual_labels == "1"); n_neg <- sum(actual_labels == "0")
baseline <- compute_financials(0, 0, n_pos, n_neg, "Baseline")
baseline_impact <- baseline$Net_Impact
fin_results$Net_Impact <- fin_results$Net_Impact - baseline_impact
baseline$Net_Impact <- 0
fin_results <- rbind(baseline, fin_results)

knitr::kable(fin_results[, c("Model", "TP", "FP", "FN", "TN", "Net_Impact", "Precision")],
             align = "lcccccc", caption = "Financial Impact vs. Baseline (Relative)")

# --- Financial bar plot ---
par(mar = c(12, 5, 3, 2))
iv <- fin_results$Net_Impact
fn <- c("Baseline", sn)
names(iv) <- fn
fc <- c("gray60", rep("#3498db",5), rep("#f39c12",5), rep("#e74c3c",2))
fc[which.max(iv)] <- "#2ecc71"
barplot(iv, col=fc, main="Net Financial Impact vs. Baseline (* = optimized threshold)",
        ylab="Net Impact ($)", las=2, cex.names=0.6)
abline(h=0, lty=2, col="gray40")
legend("topright", legend=c("Default","Optimized","Stacked"), fill=c("#3498db","#f39c12","#e74c3c"), bty="n", cex=0.8)
```

```{r}
prevalence <- mean(actual_labels == "1")
```

As observed within the Net financial impact figure, the optimized models have a much better financial impact when compared to their baselines. However, they all have the exact **same** financial impact. Looking more closely, this is because they have the **exact same** predictions! They all predict every customer as depression prone. This is a very interesting result as it indicates that the optimal financial strategy in this circumstance is simply to give every customer treatment indiscriminately. 

This unique situation is likely due to the extremely high **cost** of a misdiagnosing a depression prone customer (\$800) and the high **benefit** (\$549) of diagnosing a depression prone patient correctly. This, when combined with the skewed dataset, with `r round(prevalence * 100, 1)`% of the data points being depressed patients, meant that classifying everyone as depressed benefits from the high benefit of a true positive and the low costs of a false positive. 
