---
title: "knn_2"
author: "Daniel Myung"
date: "2026-04-13"
output: html_document
---

[← Back to Main Report](GroupFinalReport.html#knn)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this section, we apply the K-Nearest Neighbors (KNN) model to predict whether an individual has a history of mental illness. KNN is a non-parametric, instance-based classification method that makes predictions by identifying the k most similar observations in the training data and assigning the majority class among those neighbors to each new data point. 

The value of k = 15 was selected after testing multiple values, as it achieved the best Kappa score on the test set. A smaller k risks overfitting by being too sensitive to noise, while a larger k risks underfitting by over-smoothing the decision boundary. k = 15 strikes the best balance between these extremes.

```{r, cache = TRUE}
library(class)
library(caret)
train_dummy <- readRDS("train_dummy.rds")
test_dummy  <- readRDS("test_dummy.rds")

# Verification: Ensure target classes are factors
train_dummy$history_of_mental_illness_yes <- as.factor(train_dummy$history_of_mental_illness_yes)
test_dummy$history_of_mental_illness_yes  <- as.factor(test_dummy$history_of_mental_illness_yes)

# Instead of ncol-3, let's use name-based exclusion for safety
target_name_mental <- "history_of_mental_illness_yes"
# We exclude both 'yes' and 'no' dummy columns so the model doesn't cheat
exclude_cols <- c("history_of_mental_illness_yes", "history_of_mental_illness_no")

train_x <- train_dummy[, !(names(train_dummy) %in% exclude_cols)]
test_x  <- test_dummy[, !(names(test_dummy) %in% exclude_cols)]
train_y <- train_dummy[[target_name_mental]]

set.seed(12345) # For ties in KNN
knnpred_mental <- knn(train = train_x,
                      test  = test_x,
                      cl    = train_y,
                      k     = 15, prob = TRUE)

knn_model_fit <- knn3(x = train_x, 
                      y = train_y, 
                      k = 15)

# 2. SAVE THIS OBJECT
# This is what you will load into your interactive Shiny/RMD file
saveRDS(knn_model_fit, "knn_model_object.rds")

# Extract probabilities
p_mental <- attr(knnpred_mental, "prob")
# Adjusting: knn 'prob' returns the proportion of votes for the WINNING class. 
# We want the prob specifically for class '1'.
knn_prob_mental <- ifelse(knnpred_mental == "1", p_mental, 1 - p_mental)

confusionMatrix(knnpred_mental, test_dummy[[target_name_mental]], positive = "1")

saveRDS(knn_prob_mental, "knn_prob_mental_illness.rds")
saveRDS(knnpred_mental, "knn_pred_mental_illness.rds")
```

The KNN model achieves an accuracy of 67%, which falls below the No Information Rate of 69.2%, meaning a model that always predicts the majority class would actually perform better. This is confirmed by a Kappa of 0.026. The stark contrast between specificity (92.6%) and sensitivity (9.5%) reflects the class imbalance in the data: since roughly 69% of observations belong to class 0 (no history of mental illness), the model's 15 neighbors will be dominated by the majority class the vast majority of the time, pushing predictions toward 0 by default. 

As a result, the model becomes very effective at identifying individuals without a history of mental illness (true negatives) but fails to detect most of those who do have one (true positives). In practice, out of 6,365 individuals who actually had a history of mental illness, the model only correctly identified 603. 

Compared to the GLM, KNN performs notably worse. While KNN achieves a slightly higher raw accuracy (67% vs 62.5%), its Kappa of 0.026 is far below the GLM's 0.143.