---
title: "SVM"
output: html_document
---

# import data
```{r}
train <- readRDS("train_dummy.rds")

summary(train)
```


```{r}
test <- readRDS("test_dummy.rds")
```


```{r, cache=TRUE}

library(kernlab)


# radial basis rbfdot
#tanhdotmodel <- ksvm(History.of.Mental.Illness ~ ., data = train, kernel = "tanhdot")
#polydotmodel <- ksvm(History.of.Mental.Illness ~ ., data = train, kernel = "polydot")
#vanilladotmodel <- ksvm(History.of.Mental.Illness ~ ., data = train, kernel = "vanilladot")
#aplacedotmodel <- ksvm(History.of.Mental.Illness ~ ., data = train, kernel = "laplacedot")
#besseldotmodel <- ksvm(History.of.Mental.Illness ~ ., data = train, kernel = "besseldot")
#anovadotmodel <- ksvm(History.of.Mental.Illness ~ ., data = train, kernel = "anovadot")
#splinedotmodel <- ksvm(as.factor(history_of_mental_illness_yes) ~ ., data = train, kernel = "splinedot", prob.model = TRUE)
rbfdotmodel <- ksvm(as.factor(history_of_mental_illness_yes) ~ ., data = train, kernel = "rbfdot", prob.model = TRUE)


saveRDS(rbfdotmodel, file = "rbf_model.rds")
#saveRDS(tanhdotmodel, file = "tanhdot.rds")
#saveRDS(polydotmodel, file = "polydot.rds")
#saveRDS(vanilladotmodel, file = "vanilla.rds")
#saveRDS(laplacedotmodel, file = "laplace.rds")
#saveRDS(besseldotmodel, file = "bessel.rds")
#saveRDS(anovadotmodel, file = "anova.rds")
#saveRDS(splinedotmodel, file = "splinedot.rds")




#tanh_pred <- predict(tanhdotmodel, test)
#poly_pred <- predict(polydotmodel, test)
#v_pred <- predict(vanilladotmodel, test)
#l_pred <- predict(aplacedotmodel, test)
#b_pred <- predict(besseldotmodel, test)
#a_pred <- predict(anovadotmodel, test)
#s_pred <- predict(splinedotmodel, test)
rbf_pred  <- predict(rbfdotmodel, test)


#table(v_pred, tele_test$y)

## NOT A BINARY PREDICTION, so don't use confusion matrix

## this time, use confusion matrix because this is a binary prediction
library(caret)
#confusionMatrix(as.factor(tanh_pred), as.factor(test$History.of.Mental.Illness), positive = "Yes")
#confusionMatrix(as.factor(poly_pred), as.factor(test$History.of.Mental.Illness), positive = "Yes")
#confusionMatrix(as.factor(v_pred), as.factor(test$History.of.Mental.Illness), positive = "Yes")
#confusionMatrix(as.factor(l_pred), as.factor(test$History.of.Mental.Illness), positive = "Yes")
#confusionMatrix(as.factor(b_pred), as.factor(test$History.of.Mental.Illness), positive = "Yes")
#confusionMatrix(as.factor(a_pred), as.factor(test$History.of.Mental.Illness), positive = "Yes")
#confusionMatrix(as.factor(s_pred), as.factor(test$history_of_mental_illness_yes), positive = "Yes")
confusionMatrix(as.factor(rbf_pred), as.factor(test$history_of_mental_illness_yes), positive = "1")
```
```{r}
library(kernlab)

# 1. Load the model and test data
rbfdotmodel <- readRDS("rbf_model.rds")
test_dummy <- readRDS("test_dummy.rds")

# 2. Identify target columns to remove
# Because you dummied the data, you likely have TWO columns for mental illness.
# We must remove BOTH so the model doesn't see the answer.
target_cols <- c("history_of_mental_illness_no", "history_of_mental_illness_yes")
test_features <- test_dummy[, !(names(test_dummy) %in% target_cols)]

# 3. Generate probabilities 
p_rbf_matrix <- predict(rbfdotmodel, 
                           test_features, 
                           type = "probabilities")

# 4. Extract the column for the positive class
# Since your target was 'history_of_mental_illness_yes', 
# the matrix columns are likely "1" (for yes) and "0" (for no).
p_rbf_vector <- p_rbf_matrix[, "1"] 

# 5. Save the vector
saveRDS(p_rbf_vector, "rbf_prob_mental_illness.rds")
```

# summary 

After testing all the models, splinedot was the best, both in terms of Kappa and sensitivity. It got .05 kappa with 0.3 sensitivity. 

Most of the kernels performed quite terribly. Many did not hit a single prediction. The bessel model actually had a negative kappa (-0.0091). 

We hypothesize that spline dot's ability to work with non-linear connections helped it become the best kernel of the bunch. The spline kernel is known for its performance especially in sparse datasets and regression.  