---
title: "SVM"
output: html_document
---


```{r, cache=TRUE}

library(kernlab)
library(caret)

# 1. Load your pre-split Normalized/Dummy Data
train_dummy <- readRDS("train_dummy.rds")
test_dummy  <- readRDS("test_dummy.rds")

# --- DATA HEALTH CHECK ---
# Identify columns with no variance or NaNs that break SVM math
nan_cols <- sapply(train_dummy, function(x) any(is.na(x)) || var(x) == 0)

if(any(nan_cols)) {
  cat("Removing constant/problematic columns:", paste(names(nan_cols)[nan_cols], collapse=", "), "\n")
  train_dummy <- train_dummy[, !nan_cols]
  test_dummy  <- test_dummy[, !nan_cols] 
}

# 2. Define Kernel List
kernels <- c("rbfdot", "vanilladot", "polydot", "tanhdot", "laplacedot", "anovadot", "besseldot")

# 3. Training and Saving Loop
# 3. Training and Saving Loop (FORCED VERSION)
set.seed(12345)
for (k in kernels) {
  model_name <- paste0("svm_", gsub("dot", "", k))
  file_path <- paste0(model_name, ".rds")
  
  # Removed the 'if file.exists' check to force a refresh
  cat("Training (FORCED):", k, "\n")
  m <- ksvm(as.factor(history_of_mental_illness_yes) ~ ., 
            data = train_dummy, 
            kernel = k, 
            prob.model = TRUE) 
  saveRDS(m, file_path)
}

# 4. Loading and Evaluation (Categorical)
evaluation_results <- lapply(kernels, function(k) {
  model_name <- paste0("svm_", gsub("dot", "", k))
  m <- readRDS(paste0(model_name, ".rds"))
  
  # Get class predictions
  pred <- predict(m, test_dummy)
  
  # Calculate metrics focusing on class '1' (Yes)
  cm <- confusionMatrix(as.factor(pred), 
                        as.factor(test_dummy$history_of_mental_illness_yes), 
                        positive = "1")
  
  return(data.frame(
    kernel = k, 
    accuracy = cm$overall["Accuracy"], 
    kappa = cm$overall["Kappa"],
    sensitivity = cm$byClass["Sensitivity"]
  ))
})

# 5. Display Summary Table
tuning_summary <- do.call(rbind, evaluation_results)
print(tuning_summary)

# 6. Generate Probabilities with Detailed Diagnostics
p_svm_list <- lapply(kernels, function(k) {
  model_name <- paste0("svm_", gsub("dot", "", k))
  m <- readRDS(paste0(model_name, ".rds"))
  
  # Use tryCatch to prevent "atomic vector" crashes and explain failures
  prob_vector <- tryCatch({
    p <- predict(m, test_dummy, type = "probabilities")
    p[, "1"] 
    
  }, error = function(e) {
    cat("\n[FAILURE REPORT] Kernel:", k, "\n")
    
    # Diagnostic 1: Missing Internal Model
    if (is.null(m@prob.model) || length(m@prob.model) == 0) {
      cat("  - Reason: The ksvm function failed to generate a sigmoid scaling model entirely.\n")
    } 
    # Diagnostic 2: Platt Scaling Non-Convergence (The Atomic Vector Error)
    else if (is.atomic(m@prob.model[[1]])) {
      cat("  - Reason: Platt Scaling failure. The algorithm could not fit a valid sigmoid curve\n")
      cat("    to turn decision distances into probabilities. This is often caused by sparse data.\n")
    }
    cat("  - Technical Message:", conditionMessage(e), "\n")
    return(NULL)
  })
  return(prob_vector)
})
names(p_svm_list) <- kernels

# 7. Final Selection for Stacked Model
# Ensure we pick a kernel that actually has a working probability model
valid_mask <- sapply(p_svm_list, function(x) !is.null(x))
valid_summary <- tuning_summary[valid_mask, ]

if(nrow(valid_summary) > 0) {
  # Select best kernel based on Sensitivity to help our imbalanced classes
  best_kernel_name <- valid_summary$kernel[which.max(valid_summary$sensitivity)]
  
  # Save the best probability vector for the final Meta-Model script
  p_best <- p_svm_list[[best_kernel_name]]
  saveRDS(p_best, "best_svm_prob_mental_illness.rds")
}
```
```{r}
library(kernlab)
library(caret)

# 1. Load the RBF model and your test data
# Ensure test_dummy is the same version used in your SVM script
m_rbf <- readRDS("svm_rbf.rds")
test_dummy <- readRDS("test_dummy.rds")

# 1. Get RAW PROBABILITIES instead of class labels
# type = "probabilities" returns a matrix with columns "0" and "1"
prob_rbf <- predict(m_rbf, test_dummy, type = "probabilities")

# 2. Extract the probability for the "Yes" class (column "1")
p_yes <- prob_rbf[, "1"]

# 3. SET YOUR NEW THRESHOLD (e.g., 0.3)
# If the probability is > 0.3, we call it a "1"
new_threshold <- 0.30
pred_low_thresh <- ifelse(p_yes > new_threshold, "1", "0")

# 4. Convert back to factors for the Confusion Matrix
pred_low_thresh <- factor(pred_low_thresh, levels = c("0", "1"))
actual_values <- factor(test_dummy$history_of_mental_illness_yes, levels = c("0", "1"))

# 5. Run the new Confusion Matrix
cm_low_svm <- confusionMatrix(pred_low_thresh, actual_values, positive = "1")
print(cm_low_svm)
```

# summary 

After testing all the models, splinedot was the best, both in terms of Kappa and sensitivity. It got .05 kappa with 0.3 sensitivity. 

Most of the kernels performed quite terribly. Many did not hit a single prediction. The bessel model actually had a negative kappa (-0.0091). 

We hypothesize that spline dot's ability to work with non-linear connections helped it become the best kernel of the bunch. The spline kernel is known for its performance especially in sparse datasets and regression.  